Monte Carlo Expectation Maximization for RNAseq

### 2015-06-16 slides/EM/em.Rnw
mcmcem for nb

<<negative_binomial_stan, cache=TRUE>>=
negative_binomial_model = "
data {
  int<lower=1> n;
  int<lower=0> count;
  real mu_theta;
  real<lower=0> tau_theta;
  real mu_psi;
  real<lower=0> tau_psi;
}
parameters {
  real theta;
  real psi;
}
model {
  theta ~ normal(mu_theta, sqrt(tau_theta));
  psi   ~ normal(mu_psi  , sqrt(tau_psi  ));
  count ~ neg_binomial_2_log(theta, 1/psi);
}
"
nb_m = stan_model(model_code = negative_binomial_model)
@

<<negative_binomial_mcmcem, dependson=c('negative_binomial_data','negative_binomial_stan'), cache=TRUE, echo=FALSE, results='hide'>>=
M = 1000
mu_theta = 0; tau_theta=1, mu_psi = 0; mu_psi = 1 # Initial values
n_iter = 10
keep = data.frame(iteration = 1:n_iter, 
                  mu_theta = rep(NA, n_iter), 
                  tau_theta = rep(NA, n_iter), 
                  mu_psi = rep(NA,n_iter),
                  tau_psi = rep(NA,n_iter))
for (i in 1:n_iter) {
  # Run MCMC using rstan
  draws = ddply(d, .(gene), function(x) {
    s = sampling(nb_m, data = list(n = length(x$y), y=x$y, 
                                   mu_theta=mu_theta, tau_psi=tau_theta, 
                                   mu_psi=mu_psi, tau_psi=tau_psi))
    e = extract(s, c('theta','psi'))
    data.frame(theta = as.numeric(e$theta),
               psi = as.numeric(e$psi))
  })

@@

<<dependson=c('normal_hierarchical_mean_em','normal_hierarchical_mean_mcem'), echo=FALSE>>=
ggplot(keep, id.var=c("iteration"), variable.name="parameter"), 
       aes(iteration, value, color=method)) +
  geom_point() + 
  facet_wrap(~parameter, scales="free_y") +
  geom_hline(data=melt(truth, variable.name='parameter'), aes(yintercept=value))
@

>>>>

<<negative_binomial_stan, cache=TRUE>>=
negative_binomial_model = "
data {
  int<lower=1> n;
  int<lower=0> count[n];
  real mu_theta;
  real<lower=0> tau_theta;
  real mu_psi;
  real<lower=0> tau_psi;
}
parameters {
  real theta;
  real psi;
}
model {
  theta ~ normal(mu_theta, sqrt(tau_theta));
  psi   ~ normal(mu_psi  , sqrt(tau_psi  ));
  count ~ neg_binomial_2_log(theta, 1/exp(psi));
}
"
nb_m = stan_model(model_code = negative_binomial_model)
@

<<negative_binomial_mcmcem, dependson=c('negative_binomial_data','negative_binomial_stan'), cache=TRUE, echo=FALSE, results='hide'>>=
mu_theta = 0; tau_theta=1; mu_psi = 0; tau_psi = 1 # Initial values
n_iter = 100
keep = data.frame(iteration = 1:n_iter, 
                  mu_theta = rep(NA, n_iter), 
                  tau_theta = rep(NA, n_iter), 
                  mu_psi = rep(NA,n_iter),
                  tau_psi = rep(NA,n_iter))
for (i in 1:n_iter) {
  # Run MCMC using rstan
  draws = ddply(d, .(gene), function(x) {
    s = sampling(nb_m, data = list(n = length(x$y), count=x$y, 
                                   mu_theta=mu_theta, tau_theta=tau_theta, 
                                   mu_psi=mu_psi, tau_psi=tau_psi))
    e = extract(s, c('theta','psi'))
    data.frame(theta = as.numeric(e$theta),
               psi = as.numeric(e$psi))
  })

@@

<<dependson=c('normal_hierarchical_mean_em','normal_hierarchical_mean_mcem'), echo=FALSE>>=
ggplot(melt(keep, id.var=c("iteration"), variable.name="parameter"),
       aes(iteration, value)) +
  geom_point() + 
  facet_wrap(~parameter, scales="free_y") +
  geom_hline(data=melt(truth, variable.name='parameter'), aes(yintercept=value))
@
